---
title: "Visualizing the palmer penguins data with R"
author: "Meredith Fraser"
date: 2026-09-22
---

In this blog, we are going to examine the palmer penguins dataset (Gorman et al., 2014; Palmer Penguins Dataset, 2014) using R. Data were collected and made available by Dr. Kristen Gorman and the Palmer Station, Antarctica LTER. 

I am interested in visualizing which island (recorded from 3 islands in the Palmer Archipelage, Antarctica) has the longest maximum flipper length (cm) for all the Adelie penguin species measured (only species with flipper length measurements from all 3 islands).

I will first start by organizing or 'tidying' the data for our question.

```{r}
#| echo: true
#| message: false
#| warning: false
library(tidyverse)
library(palmerpenguins)

# Clean up the dataset and prep for summarizing and visualizing
penguins_clean <- penguins |>
    ## Drop `NAs` from the flipper_length_mm column (the variable we are interested in) 
    drop_na(flipper_length_mm) |>
    ## Create a new column: flipper_length_cm (converting mm to cm)
    mutate(flipper_length_cm = flipper_length_mm/100) |>
    ## Filter for only the Adelie species and select only the required columns for our question (island, flipper_length_cm)
    filter(species == "Adelie") |>
    select(island, flipper_length_cm)

# Print the head of the tidy penguins dataset
head(penguins_clean)
```

Now we can group the tidied data by island, to calculate the mean flipper length in cm for all Adelie penguins measured on that island.

```{r}
penguins_summary <- penguins_clean |>
    group_by(island) |>
    summarise(max_flipper_length_cm = max(flipper_length_cm)) 
```

I want to produce a bar plot of maximum flipper length (cm), by island for the Adelie penguin species.

```{r}
#| echo: true
#| label: fig-Adelie-penguin-flipper-R
#| table-cap: "Flipper length of species for the Adelie species"
ggplot(data = penguins_summary, mapping = aes(x = island, 
                                      y = max_flipper_length_cm,
                                     fill = island)) +
    geom_col(size = 2) +
    xlab("Penguin Species") +
    ylab("Maximum flipper length (cm)") +
    theme_bw() +
    theme(text = element_text(size = 18))
```

As you can see in Figure 1, the maximum flipper length (cm) for all Adelie penguins is similar across the three islands sampled, however an Adelie penguin on Torgersen had the longest maximum flipper length (cm), then Dream, then Biscoe.

**References**

Gorman, K. B., Williams, T. D., & Fraser, W. R. (2014). Ecological sexual dimorphism and environmental variability within a community of Antarctic penguins (genus Pygoscelis). PloS one, 9(3), e90081. https://doi.org/10.1371/journal.pone.0090081

Palmer Penguins [Dataset]. (2014). UCI Machine Learning Repository. https://doi.org/10.24432/C5R89W